---
pagetitle: "ETC3250/5250: Introduction to Machine Learning"
unitcode: "ETC3250/5250"
unitname: "Introduction to Machine Learning"
subtitle: "Decision trees"
author: "Emi Tanaka"
email: "emi.tanaka@monash.edu"
date: "Week 5"
department: "Department of Econometrics and Business Statistics"
unit-url: "iml.numbat.space"
footer: "ETC3250/5250 Week 5"
format: 
  revealjs:
    html-math-method: katex
    logo: images/monash-one-line-black-rgb.png
    slide-number: c/t
    multiplex: false
    theme: assets/monash.scss
    show-slide-number: all
    show-notes: true
    controls: true
    width: 1280
    height: 720
    css: [assets/tachyons-addon.css, assets/custom.css]
    include-after-body: "assets/after-body.html"
    auto-stretch: false
    toc: true
    toc-title: "[*Decision trees*]{.monash-blue} - table of contents"
    chalkboard:
      boardmarker-width: 5
      buttons: true
execute:
  echo: true
---


```{r, include = FALSE}
library(tidyverse)
library(gganimate)
library(tidymodels)
library(ISLR)
library(latex2exp)
library(rpart)
library(rpart.plot)
library(geomtextpath)
current_file <- knitr::current_input()
basename <- gsub(".[Rq]md$", "", current_file)

knitr::opts_chunk$set(
  fig.path = sprintf("images/%s/", basename),
  fig.width = 10,
  fig.height = 5,
  fig.align = "center",
  fig.retina = 2,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  cache.path = sprintf("cache/%s/", basename)
)

theme_set(theme_bw(base_size = 18))
```

## <br>[`r rmarkdown::metadata$unitcode`]{.monash-blue} {background-image="images/bg-01.png" #etc5523-title}

[**`r rmarkdown::metadata$unitname`**]{.monash-blue .f1}

### `r rmarkdown::metadata$subtitle`

Lecturer: *`r rmarkdown::metadata$author`*

`r rmarkdown::metadata$department`






# Classification trees {background-color="#006DAE" .mcenter}

## Classification tree


- In a **classification tree** we use a recursive two-way partition (or branches) of predictors.

. . . 

::: flex

::: {.w-40 .fragment}


```{r cancer-data}
#| echo: false
library(tidyverse)
cancer <- read_csv("https://emitanaka.org/iml/data/cancer.csv") %>% 
  mutate(diagnosis_binary = ifelse(diagnosis=="M", 1, 0),
         diagnosis = factor(diagnosis, levels = c("M", "B"))) %>% 
  janitor::clean_names()
```
```{r decision-tree}
#| echo: false
library(rpart)
library(rsample)
set.seed(2023)
cancer_split <- initial_split(cancer)
cancer_train <- training(cancer_split)
cancer_rpart <- rpart(diagnosis ~ radius_mean + concave_points_mean, 
                      data = cancer_train,
                      method = "class")
```
```{r decision-tree-plot}
#| echo: false
#| fig-width: 4.5
library(rpart.plot)
rpart.plot(cancer_rpart, box.palette = "RdGn")
```

:::

::: {.w-60 .pl3 .fragment}

```{r rpart-partitions}
#| echo: false
#| fig-width: 8
ggplot(cancer_train, aes(radius_mean, concave_points_mean)) + 
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = cancer_rpart$splits[1, "index"], ymax = Inf,
           fill = "red2", alpha = 0.2) +
  annotate("rect", xmin = -Inf, xmax = cancer_rpart$splits[2, "index"], 
           ymin = -Inf, ymax = cancer_rpart$splits[1, "index"],
           fill = "forestgreen", alpha = 0.2) +
  annotate("rect", xmin = cancer_rpart$splits[2, "index"], xmax = Inf, 
           ymin = -Inf, ymax = cancer_rpart$splits[1, "index"],
           fill = "red2", alpha = 0.2) +
  geom_texthline(yintercept = cancer_rpart$splits[1, "index"], 
                 label = "Partition 1", hjust = 0.8, vjust = -0.2,
                 linetype = "dashed", fontface = "bold") +
  
  annotate("textline", 
           x = c(cancer_rpart$splits[2, "index"], cancer_rpart$splits[2, "index"]), 
           y = c(-Inf, cancer_rpart$splits[1, "index"]), fontface = "bold",
           label = "Partition 2", vjust = 1.2, hjust = 0.05, linetype = "dashed") +
  geom_point(aes(color = diagnosis), alpha = 0.5) +
  scale_color_manual(values = c("red2", "forestgreen")) 
```


:::
:::


## <i class="fab fa-r-project"></i> Classification trees with R


- There are a number of R packages 📦 that can fit classification trees, e.g. `tree` and `rpart`. 

. . . 

- <i class="fas fa-exclamation-circle"></i> Again, be sure to read the documentation and try fitting a model to toy data to check you know how it works. 

. . . 

```{r cancer-data}
#| code-fold: true
#| code-summary: "Code for data"
```

```{r decision-tree}
```





## Interpreting classification tree plot


::: flex

::: {.w-40}

```{r decision-tree-plot}
#| fig-width: 4.5
#| code-fold: true
#| code-line-numbers: "9-11"
```

:::

::: {.w-60 .pl3}
- The rectangles are nodes that contain:
  - class identifier,
  - proportion of the reference class, and
  - percentage of observations in node.
- The bottom rectangles are **terminal nodes** (or **leaves**).
- The bold text shows the splitting rules (**branches**).
- Colors indicate class, with a darker color indicating lower _impurity_.

:::
:::


## Classifying new observations 


::: flex

::: {.w-40}

```{r decision-tree-plot}
#| fig-width: 4.5
#| code-fold: true
#| code-line-numbers: "9-11"
```

:::

::: {.w-60 .pl3}

`id` | `concave_points_mean` | `radius_mean`
---|---:|---:|
1 | 0.8 | 4.1 |
2 | 0.04 | 8.3 |
3 | 0.005 | 18.1 |


What class would above observations belong to according to this tree?

::: incremental

1. M
2. B
3. M    


:::

:::

:::




## Partition notation

::: flex
::: {.w-60}

- Suppose $s$ is the value to split in variable $j$. 
- Then we have two sets (or nodes):
  - $\color{#C8008F}{\mathcal{A}_L} = \{\{y_i, \boldsymbol{x}_i\}:x_{ij}<s\}$ and
  - $\color{#027EB6}{\mathcal{A}_R} = \{\{y_i, \boldsymbol{x}_i\}:x_{ij}\geq s\}$.
- Different choices of $j$ and $s$ encode different partitions. 
- Suppose $p_{k\mathcal{A}}$ is the proportion of observations in class $k$ in a set $\mathcal{A}$.

:::
::: {.w-40 .pl3 .f3}

```{r rpart-partition1}
#| echo: false
#| fig-width: 8
ggplot(cancer_train, aes(radius_mean, concave_points_mean)) + 
  annotate("rect", xmin = 15, xmax = Inf, 
           ymin = -Inf, ymax = Inf,
           fill = "#027EB6", alpha = 0.2) +
  annotate("rect", xmin = -Inf, xmax = 15, 
           ymin = -Inf, ymax = Inf,
           fill = "#C8008F", alpha = 0.2) +
  geom_point(aes(color = diagnosis), alpha = 0.5) +
  geom_textvline(xintercept = 15, linetype = "dashed", 
                 label = "s = 15",
                 hjust = 0.8) +
  annotate(geom = 'text', 
           x = 10, y = 0.2, 
           color = "#C8008F",
           label = TeX("$A_L$", output = 'character'),
           parse = TRUE) +
  annotate(geom = 'text', 
           x = 17, y = 0.2, 
           color = "#027EB6",
           label = TeX("$A_R$", output = 'character'),
           parse = TRUE) +
  scale_color_manual(values = c("red2", "forestgreen")) 

cancer_partition_counts <- cancer_train %>% 
    mutate(AL = ifelse(radius_mean < 15, "AL", "AR")) %>% 
    group_by(diagnosis, AL) %>% 
    count() %>% 
    pivot_wider(diagnosis, names_from = AL, values_from = n)
```


```{r rpart-partition1-counts}
#| echo: false
cancer_partition_counts %>% 
    janitor::adorn_percentages("col") %>%  
  janitor::adorn_pct_formatting() %>% 
  janitor::adorn_ns() %>% 
  mutate(diagnosis = ifelse(diagnosis == "M", "$p_{M}$", "$p_{B}$")) %>% 
  knitr::kable(escape = FALSE, 
               col.names = c("", "$\\mathcal{A}_L$", "$\\mathcal{A}_R$")) 
```

:::
:::



## Impurity metrics

::: incremental

- We need an algorithm to find the optimal $\{j^*, s^*\}$.
* Some popular [**impurity metrics**]{.monash-blue} are the: 
  - **Gini index**: $f_{Gini}(\mathcal{A})= \sum_{k = 1}^K p_{k\mathcal{A}}(1 - p_{k\mathcal{A}}) = 1 - \sum_{k = 1}^K p_{k\mathcal{A}}^2$ 
  - **entropy index**: $f_{entropy}(\mathcal{A}) = -\sum_{k = 1}^K p_{k\mathcal{A}}\log_2(p_{k\mathcal{A}})$.
* Smaller values of the impurity index means higher purity.
* An [**_overall_ impurity**]{.monash-blue} can be calculated from the weighted average:
$$\frac{|\mathcal{A}_L|}{|\mathcal{A}_L| + |\mathcal{A}_R|}f(\mathcal{A}_L) + \frac{|\mathcal{A}_R|}{|\mathcal{A}_L| + |\mathcal{A}_R|}f(\mathcal{A}_R).$$

:::

## Example calculation of Gini impurity index

```{r rpart-partition1-counts}
#| echo: false
```

```{r partition-counts-by-hand}
#| echo: false
nML <- cancer_partition_counts %>% 
  filter(diagnosis == "M") %>% 
  pull(AL)
nBL <- cancer_partition_counts %>% 
  filter(diagnosis == "B") %>% 
  pull(AL)
pL <- nML / (nML + nBL)
giniL <- round(1 - pL^2 - (1 - pL)^2, 3)

nBR <- cancer_partition_counts %>% 
  filter(diagnosis == "B") %>% 
  pull(AR)
nMR <- cancer_partition_counts %>% 
  filter(diagnosis == "M") %>% 
  pull(AR)
pR <- nMR / (nMR + nBR)
giniR <- round(1 - pR^2 - (1 - pR)^2, 3)
```

::: incremental

- $f_{Gini}(\mathcal{A}_L) = 1 - \frac{(`r nML`^2 + `r nBL`^2)}{(`r nML` + `r nBL`)^2}   \approx `r giniL`$
- $f_{Gini}(\mathcal{A}_R) = 1 - \frac{(`r nMR`^2 + `r nBR`^2)}{(`r nMR` + `r nBR`)^2}   \approx `r giniR`$
- $\text{Overall impurity} = \frac{`r nML` + `r nBL`}{`r nML` + `r nBL` + `r nMR` + `r nBR`} `r giniL` + \frac{`r nMR` + `r nBR`}{`r nML` + `r nBL` + `r nMR` + `r nBR`} `r giniR` \approx `r round(((nML + nBL) * giniL + (nMR + nBR) * giniR)/(nML + nBL + nMR + nBR), 3)`$

:::

## Search for the optimal partition 

* The goal is to find $j$ and $s$ such that it minimises the _overall impurity_:
$$\{j^*, s^*\} = \underset{j\in \{1, \dots, p\}, s \in \mathbb{R}}{\text{argmin}}~\text{Overall impurity}(\mathcal{A}_L, \mathcal{A}_R).$$


```{r data-algorithm}
#| echo: false
impurity_search <- map_dfr(c("radius_mean", "concave_points_mean"), ~{
  map_dfr(sort(unique(cancer_train[[.x]])),
          function(xval) {
            cancer_train %>%
              mutate(AL = ifelse(.data[[.x]] < xval, "AL", "AR")) %>%
              group_by(AL, diagnosis) %>%
              count() %>% 
              group_by(AL) %>% 
              mutate(p = n / sum(n)) %>% 
              summarise(gini = 1 - sum(p^2), 
                        n = sum(n)) %>% 
              mutate(weight = n / sum(n)) %>% 
              summarise(overall = sum(weight * gini)) %>% 
              mutate(xval = xval, x = .x)
          })
  })
```

::: flex
::: {.w-50}

```{r plot-algorithm1}
#| echo: false
#| fig-height: 3.1
#| fig-width: 6
ggplot(cancer_train, aes(radius_mean, concave_points_mean)) + 
  geom_point(aes(color = diagnosis), alpha = 0.5) + 
  geom_vline(aes(xintercept = xval), 
             data = impurity_search %>% 
               filter(x == "radius_mean"),
             linetype = "dashed") +
  #facet_wrap(~overall) + 
  geom_text(aes(label = sprintf("Overall impurity = %.3f", overall)), x = 20, y = 0.01,
             data = impurity_search %>% 
               filter(x == "radius_mean")) +
  transition_states(xval) +
  scale_color_manual(values = c("red2", "forestgreen")) +
  guides(color = "none")
```

:::
::: {.w-50 .pl3}
```{r plot-algorithm2}
#| echo: false
#| fig-height: 3.1
#| fig-width: 6
ggplot(cancer_train, aes(radius_mean, concave_points_mean)) + 
  geom_point(aes(color = diagnosis), alpha = 0.5) + 
  geom_hline(aes(yintercept = xval), 
             data = impurity_search %>% 
               filter(x == "concave_points_mean"),
             linetype = "dashed") +
  #facet_wrap(~overall) + 
  geom_text(aes(label = sprintf("Overall impurity = %.3f", overall)), x = 20, y = 0.01,
             data = impurity_search %>% 
               filter(x == "concave_points_mean")) +
  transition_states(xval) +
  scale_color_manual(values = c("red2", "forestgreen")) +
  guides(color = "none")
```
:::
:::




## Selected partition 

::: flex
::: {.w-50}

```{r plot-algorithm-min1}
#| echo: false
#| fig-height: 3.1
#| fig-width: 6
ggplot(cancer_train, aes(radius_mean, concave_points_mean)) + 
  geom_point(aes(color = diagnosis), alpha = 0.5) + 
  geom_vline(aes(xintercept = xval), 
             data = impurity_search %>% 
               filter(x == "radius_mean") %>% 
               filter(overall == min(overall)),
             linetype = "dashed") +
  geom_text(aes(label = sprintf("Overall impurity = %.3f", overall)), x = 20, y = 0.01,
             data = impurity_search %>% 
               filter(x == "radius_mean") %>% 
               filter(overall == min(overall))) +
  scale_color_manual(values = c("red2", "forestgreen")) +
  guides(color = "none")
```

:::
::: {.w-50 .pl3}
```{r plot-algorithm-min2}
#| echo: false
#| fig-height: 3.1
#| fig-width: 6
ggplot(cancer_train, aes(radius_mean, concave_points_mean)) + 
  geom_point(aes(color = diagnosis), alpha = 0.5) + 
  geom_hline(aes(yintercept = xval), 
             data = impurity_search %>% 
               filter(x == "concave_points_mean",
                      overall == min(overall)),
             linetype = "dashed") +
  geom_text(aes(label = sprintf("Overall impurity = %.3f", overall)), x = 20, y = 0.01,
             data = impurity_search %>% 
               filter(x == "concave_points_mean",
                      overall == min(overall))) +
  scale_color_manual(values = c("red2", "forestgreen")) +
  guides(color = "none")

min_impurity_xval <- impurity_search %>% 
  filter(overall == min(overall)) %>% 
  pull(xval)
```
:::
:::


* Selected predictor: $j^* = \texttt{concave\_points\_mean}$ 
* Selected threshold: $s^* = `r round(min_impurity_xval, 3)`$


## Search for the next optimal partition 

* We find the next $j$ and $s$ such that it minimises the overall impurity for the partitioned subset. 


```{r data-algorithm2}
#| echo: false
cancer_subset <- cancer_train %>% 
  filter(concave_points_mean <= min_impurity_xval)
impurity_search2 <- map_dfr(c("radius_mean", "concave_points_mean"), ~{
  map_dfr(sort(unique(cancer_subset[[.x]])),
          function(xval) {
            cancer_subset %>%
              mutate(AL = ifelse(.data[[.x]] < xval, "AL", "AR")) %>%
              group_by(AL, diagnosis) %>%
              count() %>% 
              group_by(AL) %>% 
              mutate(p = n / sum(n)) %>% 
              summarise(gini = 1 - sum(p^2), 
                        n = sum(n)) %>% 
              mutate(weight = n / sum(n)) %>% 
              summarise(overall = sum(weight * gini)) %>% 
              mutate(xval = xval, x = .x)
          })
  })
```

::: flex
::: {.w-50}

```{r plot-algorithm12}
#| echo: false
#| fig-height: 3.1
#| fig-width: 6
ggplot(cancer_subset, aes(radius_mean, concave_points_mean)) + 
  geom_point(aes(color = diagnosis), alpha = 0.5) + 
  geom_vline(aes(xintercept = xval), 
             data = impurity_search2 %>% 
               filter(x == "radius_mean"),
             linetype = "dashed") +
  geom_text(aes(label = sprintf("Overall impurity = %.3f", overall)), x = 16, y = 0.001,
             data = impurity_search2 %>% 
               filter(x == "radius_mean")) +
  transition_states(xval) +
  scale_color_manual(values = c("red2", "forestgreen")) +
  guides(color = "none")
```

:::
::: {.w-50 .pl3}
```{r plot-algorithm22}
#| echo: false
#| fig-height: 3.1
#| fig-width: 6
ggplot(cancer_subset, aes(radius_mean, concave_points_mean)) + 
  geom_point(aes(color = diagnosis), alpha = 0.5) + 
  geom_hline(aes(yintercept = xval), 
             data = impurity_search2 %>% 
               filter(x == "concave_points_mean"),
             linetype = "dashed") +
  geom_text(aes(label = sprintf("Overall impurity = %.3f", overall)), x = 16, y = 0.001,
             data = impurity_search2 %>% 
               filter(x == "concave_points_mean")) +
  transition_states(xval) +
  scale_color_manual(values = c("red2", "forestgreen")) +
  guides(color = "none")
```
:::
:::

::: fragment

* We repeat again for the next subset and so on ... <br>[until we reach the **stopping rule**.]{.fragment}

:::

## Stopping rules

::: incremental

- The partitioning will be recursively applied until it meets the selected stopping rule.
- Some stopping rules in `rpart.control()` include:
  - `minsplit`: the minimum number of observations in any non-terminal node.
  - `minbucket`: the minimum number of observations allowed in a terminal node.
  - `cp`: complexity parameter -- minimum difference between impurity values required to continue splitting.
  
:::

## <i class="fab fa-r-project"></i> Changing the stopping rules in `rpart`

```{r fit-change-stop}
#| code-line-numbers: "3-5"
#| fig-height: 4
#| fig-width: 7
cancer_rpart_stop <- rpart(diagnosis ~ radius_mean + concave_points_mean,
                           data = cancer_train, method = "class",
                           control = rpart.control(minbucket = 0.03 * nrow(cancer),
                                                   minsplit = 10,
                                                   cp = 0))

rpart.plot(cancer_rpart_stop, box.palette = "RdGn")
```


## Algorithm in a nutshell

::: incremental

1. Start with all observations in a single set.
2. Sort values on first variable.
3. Compute overall impurity for all possible partitions into two sets.
4. Choose the best partition on this variable.
5. Repeat 2-4 for all other variables.
6. Choose the best partition among all variables. Your data is now split into two subsets. 
7. Repeat 1-6 on each subset.
8. Stop when stopping rule is achieved. 

:::


## <i class='fab fa-r-project'></i> Tuning parameters: search space

* First we define the search space of the parameters.

```{r cv-search}
search_df <- expand_grid(minbucket = seq(10, 100, length.out = 4),
                         minsplit = seq(10, 100, length.out = 4),
                         cp = seq(0, 1, length.out = 50)) 
search_df
```

## <i class='fab fa-r-project'></i> Tuning parameters: fit and predict {.scrollable}

[<i class="fas fa-long-arrow-alt-down"></i> scroll]{.f4 .absolute .top-1 .right-1}


```{r cv-stop, cache.lazy = FALSE}
library(yardstick)
set.seed(2023)
cancer_folds <- cancer_train %>% 
  vfold_cv(v = 5) 

search_res <- cancer_folds %>% 
  mutate(search = map(splits, function(asplit) {
    search_df %>% 
      rowwise() %>% 
      # fit the model for each row 
      # different row contains unique combination of 
      # minbucket, minsplit, and cp
      mutate(fit = list(rpart(diagnosis ~ radius_mean + concave_points_mean,
                              data = training(asplit), method = "class",
                              control = rpart.control(minbucket = .data$minbucket,
                                                      minsplit = .data$minsplit,
                                                      cp = .data$cp)))) %>% 
      ungroup() %>% 
      # compute classification metric on validation fold
      mutate(metrics = map(fit, function(afit) {
        # get validation fold
        testing(asplit) %>% 
          # predict from fitted model for this validation fold
          mutate(pred = predict(afit, ., type = "class")) %>% 
          # get classification metrics
          metric_set(accuracy, bal_accuracy, kap)(., truth = diagnosis, estimate = pred)
      })) %>% 
      unnest(metrics) %>% 
      select(-c(fit, .estimator))
  })) %>% 
  unnest(search)

# summarise the data for easy view
search_res_summary <- search_res %>% 
  group_by(minbucket, minsplit, cp, .metric) %>% 
  summarise(mean = mean(.estimate),
            sd = sd(.estimate))

search_res_summary
```


## <i class='fab fa-r-project'></i> Tuning parameters: results

::: flex

::: {.w-50}

```{r vis-cv}
#| code-fold: true
#| fig-width: 7
search_res_summary_max <- search_res_summary %>% 
  group_by(.metric) %>% 
  filter(mean == max(mean))

search_res_summary %>% 
  ggplot(aes(cp, mean)) +
  geom_line(aes(color = factor(minsplit), linetype = factor(minbucket))) + 
  geom_point(data = search_res_summary_max,
             color = "red") +
  facet_grid(.metric ~ .)
```


:::

::: {.w-50 .pl3}

::: f3

```{r search-max}
#| echo: false
search_res_summary_max %>% 
  knitr::kable()
```

:::

* `minbucket` and `minsplit` doesn't seem to make much difference (for the range searched at least).
* `cp = 0` seems sufficient in this case. 


:::

:::


## <i class='fab fa-r-project'></i> Tuning `cp` with cross validation error in `rpart`

* `rpart` automatically stores the cross validation error in the resulting model object

```{r rpart-fit}
fit <- rpart(diagnosis ~ radius_mean + concave_points_mean,
             data = cancer_train, method = "class",
             control = rpart.control(cp = 0, xval = 10)) # 10 folds (default)
fit$cptable
```

* `rel error` is the in-sample error (always decreases with more split) 
* `xerror` is the cross-validation error 
* `xstd` is the standard deviation of the cross-validation error


## <i class="fas fa-database"></i> Bank marketing data {.scrollable}

[<i class="fas fa-long-arrow-alt-down"></i> scroll]{.f4 .absolute .top-1 .right-1}

- This data found [here](https://archive.ics.uci.edu/ml/datasets/Bank+Marketing) contains direct marketing campaigns of a Portuguese banking institution.
- 🎯 The goal is to predict if the client will subscribe to a term deposit (`y` = `"yes"` or `"no"`). 

```{r bank-data}
#| classes: skimr
library(tidyverse)
bank <- read_delim("https://emitanaka.org/iml/data/bank-full.csv", delim = ";")
skimr::skim(bank)
```



## <i class="fab fa-r-project"></i> Fitting and visualising the model

[<i class="fas fa-exclamation-triangle"></i> `duration` is omitted as a predictor in the model as it is computed based on the response.]{.f3}

```{r fit-bank-rpart}
#| code-line-numbers: "1|2"
#| output-location: fragment
#| fig-height: 3.5
fit_rpart <- rpart(y ~ . - duration, data = bank, method = "class")
rpart.plot(fit_rpart)
```

. . . 

* How are partitions determined for _categorical variables_?


## Partitions for categorical variables


::: flex
::: {.w-50}

- When the variable is a categorical variable, the partition of observations is based on whether it belongs to a particular class or not. 

::: {.fragment fragment-index=2}

- E.g., there are 4 classes for `poutcome` and so 4 overall impurities are calculated, one for each class vs other. 

:::

:::
::: {.w-50 .pl3 .fragment fragment-index=2}

```{r plot-categorical}
#| echo: false
#| fig-height: 4
#| fig-width: 4
cat_partition_search <- map_dfr(unique(bank$poutcome), 
        ~bank %>% 
          mutate(split = .x,
                 boutcome = ifelse(poutcome == .x, "in", "not"))) 

ggplot(cat_partition_search, aes(boutcome)) + 
  geom_bar(position = "fill", aes(fill = y), color = "black") + 
  geom_text(data = ~.x %>% 
              group_by(boutcome, split) %>% 
              summarise(n = n()),
            size = 3,
            aes(y = 1.1, label = paste0("n = ", scales::comma(n)))) +
  #facet_grid(. ~ split) +
  transition_states(split) +
  geom_label(data = ~.x %>% 
               group_by(split, boutcome) %>% 
               summarise(n = n(), 
                         y = sum(y == "yes"), 
                         p = y / n) %>%
               group_by(split) %>% 
               mutate(w = n / sum(n), 
                      gini = 1 - p^2 - (1 - p)^2) %>%
               group_by(split) %>% 
               summarise(overall = sum(w * gini)) %>% 
               mutate(boutcome = "not"),
             aes(label = str_wrap(sprintf("Overall impurity = %.3f", overall), 15), y = 0.7, x = 4), hjust = 0, nudge_x = -1) +
  ggtitle("poutcome = {closest_state}") +
  labs(y = "proportion", x = "") +
  scale_x_discrete(expand = expansion(add = c(1, 2.7))) +
  scale_y_continuous(expand = expansion(add = c(0.1, 0.3)),
                     breaks = c(0, 0.5, 1)) +
  scale_fill_manual(values = c("red2", "forestgreen")) +
  theme(legend.position = "bottom",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

:::
:::








## <i class="fas fa-check-circle"></i> Benefits of classification trees


::: incremental

- The decision rules provided by trees are **_easy to explain and follow_**.
- Trees search for optimal overall impurity independently for each predictor, therefore **_missing values in predictors can be better handled in trees_**. 
- Trees:
  - **_can handle a mix of predictor types_**, categorical and quantitative,
  - are **robust to outliers**,
  - can **capture non-linear relationships**,
  - **invariant to transformations** of predictors (as it is based on ordering).

:::

## <i class="fas fa-times-circle"></i> Limitations of classification trees


::: incremental

- Trees cannot model interaction between predictors.
- The algorithm is **greedy**!
  - This can result in a _local optimal solution_.
  - It can often _overfit_ the data (i.e. have _high variance_).
- When separation is in linear combinations of variables trees struggle to provide a good classification.

:::

# Regression tree {background-color="#006DAE" .mcenter}


## <i class="fas fa-database"></i> Insurance cost {.scrollable}

[<i class="fas fa-long-arrow-alt-down"></i> scroll]{.f4 .absolute .right-1 .top-1}

- Recall this data sourced from the [Medical Cost Personal Datasets](https://www.kaggle.com/datasets/mirichoi0218/insurance) in Week 2.

```{r insurance-data}
#| classes: skimr
library(tidyverse)
insurance <- read_csv("https://emitanaka.org/iml/data/insurance.csv")
skimr::skim(insurance)
```

## <i class="fab fa-r-project"></i> Fitting regression tree with R 

```{r reg-tree}
fit <- rpart(charges ~ ., data = insurance, method = "anova")
rpart.plot(fit)
```


## Interpreting regression tree plot


::: flex

::: {.w-40}

```{r reg-tree-plot}
#| fig-width: 4.5
#| echo: false
rpart.plot(fit)
```

:::

::: {.w-60 .pl3}
- The nodes contain:
  - **average of the response**, and
  - percentage of observations in node.
- The bold text shows the splitting rules (branches).
- Colors indicate class, with a darker color indicating lower impurity.
- The **prediction is the average of the response in the terminal node**.

:::
:::

## Partition based on a binary predictor



```{r insurance-impurity}
#| echo: false
insurance_partition1 <- insurance %>% 
               group_by(smoker) %>% 
               summarise(avg = mean(charges),
                         RSS = sum((charges - mean(charges))^2))

overall1 <- sum(insurance_partition1$RSS)
```


::: flex
::: {.w-60 .incremental}

- For a binary variable, $\mathcal{A}_L$ and $\mathcal{A}_R$ is based on its class. 
- Then find the average responses, $\bar{y}_L$ and $\bar{y}_R$, for $\mathcal{A}_L$ and $\mathcal{A}_R$, respectively.
- We can use the **residual sum of squares**, e.g. $\text{RSS}(\mathcal{A}_L) = \sum_{i \in \mathcal{A}_L}(y_i - \bar{y}_L)^2$, to measure the impurity. 
- The **overall impurity** is $\text{RSS}(\mathcal{A}_L) + \text{RSS}(\mathcal{A}_R) = `r round(overall1/1000/1000/1000, 1)`\text{B}$.

:::
::: {.w-40 .pl3}

```{r plot-insurance}
#| echo: false
#| fig-width: 5
#| fig-height: 6
ggplot(insurance, aes(bmi, age)) +
  geom_point(aes(color = charges)) +
  colorspace::scale_color_continuous_sequential(palette = "Inferno", trans = "log10") +
  facet_wrap(~smoker, labeller = label_both, ncol = 1) +
  geom_label(aes(label = paste(sprintf("Average charge = %.1fK", avg/1000), sprintf("RSS = %.1fB", RSS/1000/1000/1000), sep = "\n"), x = 20, y = 58, hjust = 0),
             fontface = "bold",
             data = insurance_partition1)
```

:::
:::


## Partition based on a quantitative predictor

- Like for classification trees, a quantitative predictor is partitioned based on the split value. 
- The overall impurity is based on a combined impurity measure for quantitative response (RSS here).

```{r insurance-data-algorithm}
#| echo: false
insurance_impurity_search <- map_dfr(c("age", "bmi"), ~{
  map_dfr(sort(unique(insurance[[.x]])),
          function(xval) {
            insurance %>%
              mutate(AL = ifelse(.data[[.x]] < xval, "AL", "AR")) %>%
              group_by(AL) %>%
              summarise(RSS = sum((charges - mean(charges))^2)) %>% 
              ungroup() %>% 
              summarise(overall = sum(RSS)) %>% 
              mutate(xval = xval, x = .x)
          })
  })
```

::: flex
::: {.w-50}

```{r insurance-plot-algorithm1}
#| echo: false
#| fig-height: 3.1
#| fig-width: 6
ggplot(insurance, aes(bmi, age)) + 
  geom_point(aes(color = charges), alpha = 0.5) + 
  geom_vline(aes(xintercept = xval), 
             data = insurance_impurity_search %>% 
               filter(x == "bmi"),
             linetype = "dashed") +
  #facet_wrap(~overall) + 
  geom_text(aes(label = sprintf("Overall impurity = %.1fB", overall/1000/1000/1000)), 
            x = 40, y = 58, hjust = 0,
            data = insurance_impurity_search %>% 
               filter(x == "bmi"),
            fontface = "bold") +
  transition_states(xval) +
  colorspace::scale_color_continuous_sequential(palette = "Inferno", trans = "log10") +
  guides(color = "none") 
```

:::
::: {.w-50 .pl3}
```{r insurance-plot-algorithm2}
#| echo: false
#| fig-height: 3.1
#| fig-width: 6
ggplot(insurance, aes(bmi, age)) + 
  geom_point(aes(color = charges), alpha = 0.5) + 
  geom_hline(aes(yintercept = xval), 
             data = insurance_impurity_search %>% 
               filter(x == "age"),
             linetype = "dashed") +
  #facet_wrap(~overall) + 
  geom_text(aes(label = sprintf("Overall impurity = %.1fB", overall/1000/1000/1000)), 
            x = 40, y = 58, hjust = 0,
             data = insurance_impurity_search %>% 
               filter(x == "age"),
            fontface = "bold") +
  transition_states(xval) +
  colorspace::scale_color_continuous_sequential(palette = "Inferno", trans = "log10") +
  guides(color = "none") 
```
:::
:::





## <i class='fab fa-r-project'></i> Pruning

- We can grow a tree, $T_0$, and then **prune** it to a shorter tree ($T_1$). 


::: flex

::: {.w-50 .mcenter .fragment}

$T_0$

```{r full-tree}
T0 <- rpart(charges ~ ., 
            data = insurance, 
            method = "anova", 
            cp = 0)
rpart.plot(T0)
```

:::

::: {.w-50 .pl3 .mcenter .fragment}

$T_1$

```{r pruned-tree}
#| output-location: fragment
T1 <- prune(T0, cp = 0.01)
rpart.plot(T1)
```


:::

:::

## <i class='fab fa-r-project'></i> Pruning to optimal `cp` based on `xerror` {.scrollable}

[<i class="fas fa-long-arrow-alt-down"></i> scroll]{.f4 .absolute .top-1 .right-1}


* We can select optimal `cp` based on `xerror`



```{r cptable}
T0$cptable %>% as_tibble()
optimal_cp <- T0$cptable %>% 
              as.data.frame() %>% 
              filter(xerror == min(xerror)) %>% 
              # if multiple optimal points, then select one
              slice(1) %>% 
              pull(CP)
optimal_cp

T2 <- prune(T0, cp = optimal_cp)
rpart.plot(T2)
```










## Decision tree algorithm

- **Top-down**: it begins at the top of the tree (all observations belong to a single region) and then successively splits the predictor space; each split is indicated via two new branches further down on the tree.
- **Greedy**: at each step of the tree-building process, the best split is made at that particular step, rather than looking ahead and picking a split that will lead to a better tree in some future step.


# <i class="fas fa-key"></i> Takeaway {background-color="#006DAE"}


- Tree-based models consist of one or more of nested conditions for the predictors that partition the data. 
- **Decision trees** can be used for both regression and classification problems.
- Decision trees used for:
  - classification problems is referred to as **classification trees**, and
  - regression problems is referred to as **regression trees**.
